Do Alternate Wetting and Drying Irrigation Technology and Nitrogen Rates Affect Rice Sheath Blight?

### 2018-04-14 analysis/paper/paper.Rmd
Major edits to text and output
- Add figures 1-3 of disease progress over time

- Add more text to experimental design and results

- Add new reference for Nancy's 1996 paper

- Add csl for EJPP

- Add Word template for knitting Word, but waiting on PR to be accepted in `rrtools` with new Springer template that we can use

---
title: Do Alternate Wetting and Drying Irrigation Technology and Nitrogen Rates Affect Rice Sheath Blight?
author: |
  | Adam H. Sparks^[Corresponding author: Adam.Sparks@usq.edu.au]  \textsuperscript{1\dag}, Nancy P. Castilla \textsuperscript{1}, Bjorn Ole Sander \textsuperscript{1}
  | \textsuperscript{1}International Rice Research Institute, Crop and Environmental Sciences Division, Los Baños, Laguna, Philippines, 4031
  | \textsuperscript{\dag} Current Address: University of Southern Queensland, Centre for Crop Health, West Street, Tooowoomba, Queensland, Australia, 4350
abstract: |
  Water and nitrogen management play vital roles in rice production. However, the mismanagement of these two management practices may trigger sheath blight of rice, caused by _Rhizoctonia solani_, which is favored by wet conditions, high relative humidity, and high nitrogen fertilizer levels. To understand how different combinations of water and nitrogen management affect sheath blight epidemics, we conducted two separate split-plot experiments with a water saving (alternate wetting and drying) regime and traditional flood irrigation regime combined with differing nitrogen treatments in the dry seasons of 2015 and 2016. Disease was scored in the same way in both experiments using a sheath blight assessment scale for field evaluation developed at the International Rice Research Institute to assess the severity on infected sheaths and leaves while sheath blight incidence on tillers were counted per hill. We were unable to detect any difference in the incidence of tiller sheath blight due to irrigation, tiller and leaf sheath blight did differ significantly by irrigation treament but leaf sheath blight severity did not. Our findings suggests that farmers can adopt water saving technologies without risking increased sheath blight incidence. We suggest that further cross-cutting research in this area is warranted.

bibliography: references.bib
output:
  bookdown::word_document2:
    toc: no
    fig_caption: yes

always_allow_html: yes

header-includes:
  - \usepackage{endfloat}
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}
  - \linenumbers
---

```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(fig.pos = "p")
knitr::opts_chunk$set(out.width = "100%", dpi = 300)

library(bookdown)
library(knitr)
library(kableExtra)
```

# Introduction

Alternate wetting and drying (AWD) is an irrigation technique for irrigated rice (\emph{Oryza sativa L.}) developed by the International Rice Research Institute (IRRI) and its partners that saves about 15-40% of irrigation water [@Bouman2001; @Feng2007]. In AWD rice, fields are exposed to several dry phases during the growth period without exposing the plants to water stress. In order to avoid yield decline under AWD “safe” thresholds have been developed. Under safe AWD irrigation water is applied when the field water level reaches 15cm below the soil surface [@Richards2014]. Fields are furthermore kept flooded during the flowering period. As an added benefit to saving water, AWD also reduces greenhouse gas (GHG) emissions of rice fields, which are a substantial factor in the GHG budget of rice producing countries, by around 50% [@Yan2005; @Sander2015].

The AWD technology has been identified as promising climate smart practice for different rice growing regions that can stabilize rice production in water scarce areas as well as help reduce the carbon footprint of rice production. Various countries, \textit{e.g.}, Bangladesh, Vietnam, Thailand and the Philip pines, plan to widely apply AWD to local rice production [@MOEF2015]. However, a change in water regime in rice fields on large scale might encompass different other effects, for example related to plant health.

We therefore established field experiments in order to determine what effects AWD could have on sheath blight disease (\emph{Rhizoctonia solani} Kühn), anastamosis group 1 [(teleomorph: \emph{Thanatephorus cucumeris} (A.B. Frank) Donk.] of rice.

@@

# Materials and Methods

## Crop establishment

@@

## Inoculum

@@

## Experimental Design

@@

**2015 Dry Season Experiment** The main plot size was 12m x 12m (144 sq m), with a sub-plot size of 5m x 5m (25 sq m). Replication size was 12m x 24m (288 sq m) with a buffer of 1m per sub-plot for a whole experiment size of 1,152 sq m.

Irrigation in AWD plots was determined by the water level in plots, \emph{i.e.}, when the water level reached 15cm below the soil surface irrigation water was applied to a level of 5cm. In CF plots a standing water layer of 3-5cm was maintained throughout the growing season.

@@

**2016 Dry Season Experiment** In 2016 dry season the plot size was increased and due to these changes, the replicate sizes are are not equal as necessitated by the use of a larger area for the experiment. The main plot sizes were: Block 1 (B1) 21m x 20.5m (412.5 sq m) and Block 2 (B2) 20.25m x 21.6m (437.4 sq m). The sub-plot sizes were B1 21m x 10.25m (215.25 sq m), B2 20.25m x 10.8m (218.7 sq m). The replication sizes were B1 - 42m x 20.5m (861 sq m) and B2 - 40.5m x 21.6m (874.8 sq m). A buffer 0.5m per sub-plot was used and the overall experiment size was 3471.6 sq m.

@@

## Data Collection and Analysis

Disease scoring was the same in both experiments using the same sheath blight assessment scale for field evaluation developed at IRRI \ref{figure1}. Two sample areas per plot (1m x 1m) were assessed. For nine hills per sample, the number of tillers per hill and number of tillers with sheath blight (incidence) were measured. Tiller sheath blight severity was measured for four tillers per hill and six leaves tiller using the rating scale. Five disease assessments were made in the 2015 experiment and four disease assessments were made in the 2016 experiment, respectively.

![IRRI sheath blight field severity rating scale for rice leaf tissue where, 0 - 0%, 1 - 1%, 2 - 5%,
3 - 15%, 4 - 50%, 5 - >50% severity. \label{figure1}](./Figures/Figure 1 IRRI Sheath Blight Severity Assessment.png)

Disease incidence and severity were converted to area under the disease progress stairs (AUDPS) [@simko2012area]. As most of the data's residuals did not meet assumptions for normality, the analysis was carried out using multivariate generalised linear mixed models implemented in the MCMCglmm package [@MCMCglmm2010] in the R programming environment [@R2018].

# Results

### Tiller Sheath Blight Incidence

In 2015 both of the nitrogen treatements, N100 and N120, were significantly different when compared with the control N0 treatement. However, water management was not significantly different.

In 2016 the nitrogen treatment N180, was significantly different than the control N60 treatment. As in the 2015 study, water management did not significantly differ.

### Tiller Sheath Blight Severity

@@

In 2016 the N180 treatement was significantly different from the N60 treatment. The AWD water management was also significantly different from the control flooding treatment.

### Leaf Sheath Blight Severity

@@

# Discussion

# Acknowledgments

@@

# References 
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

@@

### Colophon

>>>>

---
title: Do Alternate Wetting and Drying Irrigation Technology and Nitrogen Rates Affect Rice Sheath Blight?
author: |
  | Adam H. Sparks^[Corresponding author: Adam.Sparks@usq.edu.au]  \textsuperscript{1\dag}, Nancy P. Castilla \textsuperscript{1}, Bjorn Ole Sander \textsuperscript{1}
  | \textsuperscript{1}International Rice Research Institute, Crop and Environmental Sciences Division, Los Baños, Laguna, Philippines, 4031
  | \textsuperscript{\dag} Current Address: University of Southern Queensland, Centre for Crop Health, West Street, Tooowoomba, Queensland, Australia, 4350
csl: "../templates/european-journal-of-plant-pathology.csl" # Insert path for the bib-style
abstract: |
  Water and nitrogen management play vital roles in rice production. However, the mismanagement of these two management practices may trigger sheath blight of rice, caused by _Rhizoctonia solani_, which is favored by wet conditions, high relative humidity, and high nitrogen fertilizer levels. To understand how different combinations of water and nitrogen management affect sheath blight epidemics, we conducted two separate split-plot experiments with a water saving (alternate wetting and drying) regime and traditional flood irrigation regime combined with differing nitrogen treatments in the dry seasons of 2015 and 2016. Disease was scored in the same way in both experiments using a sheath blight assessment scale for field evaluation developed at the International Rice Research Institute to assess the severity on infected sheaths and leaves while sheath blight incidence on tillers were counted per hill. We were unable to detect any difference in the incidence of tiller sheath blight due to irrigation, tiller and leaf sheath blight did differ significantly by irrigation treament but leaf sheath blight severity did not. Our findings suggests that farmers can adopt water saving technologies without risking increased sheath blight incidence. We suggest that further cross-cutting research in this area is warranted.
bibliography: references.bib
output:
  bookdown::word_document2:
    toc: no
    fig_caption: yes
    reference_docx: "../templates/template.docx"
always_allow_html: yes
header-includes:
   \usepackage{endfloat}
   \usepackage{setspace}\doublespacing
   \usepackage{lineno}
   \linenumbers
---

```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(fig.pos = "p")
knitr::opts_chunk$set(out.width = "100%", dpi = 300)
knitr::opts_chunk$set(eval = TRUE)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)

library(bookdown)
library(knitr)
library(kableExtra)
library(rice.awd.pests)
library(ggpubr)
```

# Introduction {-}

Alternate wetting and drying (AWD) is an irrigation technique for irrigated rice (*Oryza sativa* L.) developed by the International Rice Research Institute (IRRI) and its partners that saves about 15-40% of irrigation water [@Bouman2001; @Feng2007]. In AWD rice, fields are exposed to several dry phases during the growth period without exposing the plants to water stress. In order to avoid yield decline under AWD “safe” thresholds have been developed. Under safe AWD irrigation water is applied when the field water level reaches 15cm below the soil surface [@Richards2014]. Fields are furthermore kept flooded during the flowering period. As an added benefit to saving water, AWD also reduces greenhouse gas (GHG) emissions of rice fields, which are a substantial factor in the GHG budget of rice producing countries, by around 50% [@Yan2005; @Sander2015].

The AWD technology has been identified as promising climate smart practice for different rice growing regions that can stabilize rice production in water scarce areas as well as help reduce the carbon footprint of rice production. Various countries, *e.g.*, Bangladesh, Vietnam, Thailand and the Philip pines, plan to widely apply AWD to local rice production [@MOEF2015]. However, a change in water regime in rice fields on large scale might encompass different other effects, for example related to plant health.

Nitrogen is known to affect sheath blight [@Castilla1996]...

We therefore established field experiments in order to determine what effects AWD could have on sheath blight disease (*Rhizoctonia solani* Kühn), anastamosis group 1 [(teleomorph: *Thanatephorus cucumeris* (A.B. Frank) Donk.] of rice.

@@

# Materials and Methods {-}

## Crop establishment {-}

@@

## Inoculum {-}

@@

## Experimental Design {-}

@@

### 2015 Dry Season Experiment {-}

The main plot size was 12m x 12m (144 sq m), with a sub-plot size of 5m x 5m (25 sq m). Replication size was 12m x 24m (288 sq m) with a buffer of 1m per sub-plot for a whole experiment size of 1,152 sq m.

Irrigation in AWD plots was determined by the water level in plots, *i.e.*, when the water level reached 15cm below the soil surface irrigation water was applied to a level of 5cm. In CF plots a standing water layer of 3-5cm was maintained throughout the growing season.

@@

### 2016 Dry Season Experiment {-}

In 2016 dry season the plot size was increased and due to these changes, the replicate sizes are are not equal as necessitated by the use of a larger area for the experiment. The main plot sizes were: Block 1 (B1) 21m x 20.5m (412.5 sq m) and Block 2 (B2) 20.25m x 21.6m (437.4 sq m). The sub-plot sizes were B1 21m x 10.25m (215.25 sq m), B2 20.25m x 10.8m (218.7 sq m). The replication sizes were B1 - 42m x 20.5m (861 sq m) and B2 - 40.5m x 21.6m (874.8 sq m). A buffer 0.5m per sub-plot was used and the overall experiment size was 3471.6 sq m.

@@

## Data Collection and Analysis {-}

Disease scoring was the same in both experiments using the same sheath blight disease assessment scale for field evaluation developed at IRRI where 0 - no disease; 1 - trace to 1% severity; 2 - 1 to 5%; 3 - 5 to 15%; 4 - 15 to 50%; 5 - 50 to 100%. Where severity was considered to be the amount of leaf or sheath tissue covered by sheath blight lesions. Two sample areas per plot (1m x 1m) were assessed. For nine hills per sample, the number of tillers per hill and number of tillers with sheath blight (incidence) were measured. Tiller sheath blight severity was measured for four tillers per hill and six leaves tiller using the rating scale. Five disease assessments were made in the 2015 experiment and four disease assessments were made in the 2016 experiment, respectively.

Disease severity values were converted to the midpoint percentage value of the disease assessment scale. Area under the disease progress stairs (AUDPS) [@simko2012area] was calculated for both disease incidence and severity. As most of the data's residuals did not meet assumptions for normality, the analysis was carried out using multivariate generalised linear mixed models implemented in the `MCMCglmm` package [@MCMCglmm2010] in the R programming environment [@R2018].

# Results {-}

## Tiller Sheath Blight Incidence {-}

In 2015 the incidence of tiller sheath blight remained low throughout the growing season only reaching 15% in 2015 in the N120/flooded treatment (Figure \@ref(fig:figure-1)). The nitrogen treatments, N100 and N120, were both significantly different when compared with the control N0 treatment. However, water management was not significantly different.

```{r figure-1, fig.cap="Tiller sheath blight incidence progress was measured in the 2015 experiment at five assessment points and 2016 experiment at four assessment points during the dry season both years. Main plot irrigation treaments were permanently flooded (FLD) and alternate wetting and drying (AWD). Five different nitrogen treatments were applied depending on the experiment. In the 2015 experiment three nitrogen rate treatements were applied: no N (N0), 100 units of N (N100) and 180 units of N (N180). In the 2016 experiment, two nitrogen rate treatments were applied: 60 units of N (N60) and 180 units of N (N180)."}

ggplot(RAW_data,
       aes(x = ASMT,
           y = TShB_incidence,
           linetype = NRTE,
           shape = NRTE)) +
  stat_summary(fun.y = "mean",
               geom = "line",
               aes(group = TRT)) +
  stat_summary(fun.y = "mean",
               geom = "point",
               aes(group = TRT)) +
  scale_y_continuous(labels = function(x) paste0(x*100), limits = c(0, 1)) +
  scale_shape_discrete(labels = c("N0", "N100", "N120", "N60", "N180")) +
  scale_linetype_discrete(labels = c("N0", "N100", "N120", "N60", "N180")) +
  ylab("Tiller Sheath Blight Incidence (%)") +
  xlab("Assessment") +
  facet_grid(WMGT ~ YEAR) +
  labs(linetype = "Nitrogen treatment",
       shape = "Nitrogen treatment") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "black")) +
  theme(legend.position = "bottom")
```

The changes to the inoculation methodology resulted in a much higher rate of infection in 2016, with the N180/flooded treatment reaching a maximum value of 98% incidence at the third observation (Figure \@ref(fig:figure-1)). The nitrogen treatment N180, was significantly different than the control N60 treatment. As in the 2015 study, water management did not significantly differ.

## Tiller Sheath Blight Severity {-}

@@

In 2016 the N180 treatment was significantly different from the N60 treatment. The AWD water management was also significantly different from the control flooding treatment.

```{r figure-2, fig.cap="Tiller sheath blight severity progress was measured in the 2015 experiment at five assessment points and 2016 experiment at four assessment points during the dry season both years. Main plot irrigation treaments were permanently flooded (FLD) and alternate wetting and drying (AWD). Five different nitrogen treatments were applied depending on the experiment. In the 2015 experiment three nitrogen rate treatements were applied: no N (N0), 100 units of N (N100) and 180 units of N (N180). In the 2016 experiment, two nitrogen rate treatments were applied: 60 units of N (N60) and 180 units of N (N180)."}

ggplot(RAW_data,
       aes(x = ASMT,
           y = PERC_TIL_ShB,
           linetype = NRTE,
           shape = NRTE)) +
  stat_summary(fun.y = "mean",
               geom = "line",
               aes(group = TRT)) +
  stat_summary(fun.y = "mean",
               geom = "point",
               aes(group = TRT)) +
  scale_shape_discrete(labels = c("N0", "N100", "N120", "N60", "N180")) +
  scale_linetype_discrete(labels = c("N0", "N100", "N120", "N60", "N180")) +
  ylab("Tiller Sheath Blight Severity (%)") +
  xlab("Assessment") +
  facet_grid(WMGT ~ YEAR) +
  labs(linetype = "Nitrogen treatment",
       shape = "Nitrogen treatment") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "black")) +
  theme(legend.position = "bottom")
```

## Leaf Sheath Blight Severity {-}

@@

```{r figure-3, fig.cap="Leaf sheath blight severity progress was measured in the 2015 experiment at five assessment points and 2016 experiment at four assessment points during the dry season in both years. Main plot irrigation treaments were permanently flooded (FLD) and alternate wetting and drying (AWD). Five different nitrogen treatments were applied depending on the experiment. In the 2015 experiment three nitrogen rate treatements were applied: no N (N0), 100 units of N (N100) and 180 units of N (N180). In the 2016 experiment, two nitrogen rate treatments were applied: 60 units of N (N60) and 180 units of N (N180)."}

ggplot(RAW_data,
       aes(x = ASMT,
           y = PERC_LEAF_ShB,
           linetype = NRTE,
           shape = NRTE)) +
  stat_summary(fun.y = "mean",
               geom = "line",
               aes(group = TRT)) +
  stat_summary(fun.y = "mean",
               geom = "point",
               aes(group = TRT)) +
  scale_shape_discrete(labels = c("N0", "N100", "N120", "N60", "N180")) +
  scale_linetype_discrete(labels = c("N0", "N100", "N120", "N60", "N180")) +
  ylab("Leaf Sheath Blight Severity (%)") +
  xlab("Assessment") +
  facet_grid(WMGT ~ YEAR) +
  labs(linetype = "Nitrogen treatment",
       shape = "Nitrogen treatment") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "black")) +
  theme(legend.position = "bottom")
```

# Discussion {-}

# Notes {-}

## Acknowledgments {-}

@@

# References {-}
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

@@

### Colophon {-}
